Inequality in socioemotional skills: a cross-cohort comparison

### 2020-03-14 codes/R/paper/text/paper_jpube_1introlit.tex
update text

[\textbf{ADD REFS}]

Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle. Human capital is also important for various determinants of individual well-being, ranging from health to life satisfaction.  In recent years, the process of human capital accumulation has received considerable attention. There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development. It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and longlasting role.

And yet, there are still large gaps in our knowledge of the process of human capital development. These gaps are partly driven by the scarcity of high quality longitudinal data measuring the evolution over the life cycle of different dimension of human capital. Moreover, there is a lack of consensus on the best measures and on the tools to collect high quality data. As a consequence, even when data are available in different contexts, their comparability is problematic.

@@

We proceed in four steps. First, we construct a novel scale of childhood behavioural traits from two validated instruments and assess its comparability across cohorts. By performing an exploratory factor analysis, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts. Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination, and the latter relating to their ability to engage in interpersonal activities. Importantly, for the first time in economics, we study the comparability of the measures in the two cohorts. In particular, we test for \emph{measurement invariance} of the items we use to estimate the latent factors. Intuitively, if one assumes that a set of measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level. If one uses a given measure as the relevant metric for the relevant factor, its saliency will determine the scale of the factor, while some other parameters, which could be driven by the difficulty of a given test or the social norms and attitudes towards a certain type of behaviour, determine the \emph{average level} of the factor. Comparability of estimated factors across different groups (such as different cohorts) assumes  that both the parameters that determine the saliency of a given set of measures and the level of the factors do not vary across groups. We find that, for the measures we use and for both factors, we cannot reject measurement invariance for the saliency parameters. However, we strongly reject measurement invariance for the level parameters. These results imply that while we can compare the inequality in skills across the two cohorts, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts. While this result hinders a comparison in the level of skills, it is of interest per se to find that mothers of children born in England thirty years apart assess behaviours differently, so that differences in the raw scales cannot be unequivocally interpreted as differences in the underlying skills. We believe this is an important finding which deserves a greater degree of attention in the economic literature.

>>>>

Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle \citep{hanushek2008role}. Human capital is also important for various determinants of individual well-being, ranging from life satisfaction to health \citep{conti2019developmental}. In recent years, the process of human capital accumulation has received considerable attention \citep{almond2018childhood}. There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development. It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and longlasting role \citep{Cunha2010}.

And yet, there are still large gaps in our knowledge of the process of human capital development. These gaps are partly driven by the scarcity of high quality longitudinal data measuring the evolution over the life cycle of different dimension of human capital. Moreover, there is a lack of consensus on the best measures and on the tools to collect high quality data. As a consequence, even when data are available in different contexts, their comparability is problematic \citep{richter2019early}.

@@

We proceed in four steps. First, we construct a novel scale of childhood behavioural traits from two validated instruments and assess its comparability across cohorts. By performing exploratory and multiple-group factor analyses, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts. Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination, and the latter relating to their ability to engage in interpersonal activities. Importantly, for the first time in economics, we study the comparability of the measures in the two cohorts. In particular, we test for \emph{measurement invariance} of the items we use to estimate the latent factors. Intuitively, if one assumes that a set of measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level. If one uses a given measure as the relevant metric for the relevant factor, its saliency will determine the scale of the factor, while some other parameters, which could be driven by the difficulty of a given test or the social norms and attitudes towards a certain type of behaviour, determine the \emph{average level} of the factor. Comparability of estimated factors across different groups (such as different cohorts) assumes  that both the parameters that determine the saliency of a given set of measures and the level of the factors do not vary across groups. We find that, for the measures we use and for both factors, we cannot reject measurement invariance for the saliency parameters. However, we strongly reject measurement invariance for the level parameters. These results imply that while we can compare the inequality in skills across the two cohorts, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts. While this result hinders a comparison in the level of skills, it is of interest per se to find that mothers of children born in England thirty years apart assess behaviours differently, so that differences in the raw scales cannot be unequivocally interpreted as differences in the underlying skills. We believe this is an important finding which deserves a greater degree of attention in the economic literature.

### 2020-03-14 codes/R/paper/text/paper_jpube_2datamethods.tex
update text

We use exploratory factor analysis (EFA) to assess the factor structure of our new scale, composed of 11 items of the Rutter scale in the BCS and the corresponding items of the SDQ in the MCS.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and recently adopted by the economics literature. The results are displayed in \autoref{tab:numfac}. As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors for the SDQ items in the MCS. In our analysis, we adopt two factors, which is the average number of factors across all methods for each cohort and gender. This choice is also consistent with the child psychology literature cited above, and allows us to work with the same number of factors in the two cohorts.

The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very similar. Following previous research, we name the first dimension \emph{Externalising skills} (EXT, indicating low scores on the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, indicating low scores on the items worried, fearful, solitary, unhappy, and aches).\footnote{Internalising and externalising dimensions emerge from the exploratory step on our novel 11-item scale. Appendix \autoref{asec:fullefa} performs the same exploratory steps on the full set of Rutter items in BCS and SDQ items in MCS. It confirms that the items we select for our subscale have a broadly consistent covariance structure even when factor-analysed with the others in their original scales. Appendix \autoref{asec:no511} considers the robustness of our results to the exclusion of the items of the scale that perform most poorly.}

@@

Equipped with the factor structure inferred in the previous section, we specify a multiple-group factor analysis model to formally quantify the strength of the relationship between the observed items in our scale and the two latent socio-emotional skills. We specify two groups of children $c=\{BCS,MCS\}$, corresponding to the two cohorts. Each individual child is denoted by $j=1\dots N_c$, where $N_c$ is the number of children in cohort $c$. For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$ corresponding to the eleven maternal reports in \autoref{tab:scalecomp}. We assume that each child is characterised by a latent bidimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$.

@@

We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}. We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$} This choice of using a dedicated measurement system is consistent with the standard rule of thumb for selecting items based on the magnitude of their loadings, which is indicated as 0.32 by \citet{Tabachnick2001}.

@@

Any comparison between socio-emotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts. In other words, the items in our new scale must measure externalising and internalising socio-emotional skills in the same way in the BCS and MCS data. This property is denominated measurement invariance (MI) \citep{Vandenberg2000a,Putnick2016}.

@@

In the case where the available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).

Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood. In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$). This makes identification reliant on more stringent normalisations. A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}. New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed. In light of this, they propose models that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts. A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.

@@

In any case, we present a range of fit indices to provide a more complete assessment of measurement invariance. The fit of each nested model is compared in Panel A of \autoref{tab:fit}. The overall fit of the configural model is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$. As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels. The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs. In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance. Invariance of loadings and thresholds across cohorts implies that the items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.

\textbf{[ADD NEW RESULTS]}

>>>>

We use exploratory factor analysis (EFA) to assess the factor structure of our new scale, composed of 11 items of the Rutter scale in the BCS and the corresponding items of the SDQ in the MCS.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and recently adopted by the economics literature. The results are displayed in \autoref{tab:numfac}. As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors for the SDQ items in the MCS.

Given the test results, we perform a series of exploratory factor analyses, assuming a one-, two- or three-factor structure, respectively. The results for the 1-factor solution, reported in \autoref{tab:load_add}, show relatively similar loadings for both males and females across the two cohorts, of slightly bigger magnitude for the last four items in the MCS than in the BCS; thus, we retain the 1-factor solution for the measurement invariance analysis, in the first instance. The results for the 3-factor solution, instead, also reported in \autoref{tab:load_add}, show a less homogeneous picture:\footnote{We do not perform the EFA with 3 factors for males because this solution is never chosen by any test for the number of factors for the MCS, see \autoref{tab:numfac}.} while the magnitude of the loadings is relatively similar across the two cohorts for the first factor, items 3 and 5 only load on the second factor for the MCS, not for the BCS; more importantly, the EFA clearly shows that the third factor only loads on one single item (item number 9, ``solitary'') for both cohorts. Given that a one-item factor implies that the item perfectly proxies for the factor, we are not able to test for measurement invariance in this case. Hence, the 3-factor solution is not supported by our EFA results. Last, the two-factor EFA is shown in \autoref{tab:load} and delivers a neat and sensible separation between items: similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very similar. Following previous research, we name the first dimension \emph{Externalising skills} (EXT, indicating low scores on the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, indicating low scores on the items worried, fearful, solitary, unhappy, and aches).\footnote{Internalising and externalising dimensions emerge from the exploratory step on our novel 11-item scale. Appendix \autoref{asec:fullefa} performs the same exploratory steps on the full set of Rutter items in BCS and SDQ items in MCS. It confirms that the items we select for our subscale have a broadly consistent covariance structure even when factor-analysed with the others in their original scales. Appendix \autoref{asec:no511} considers the robustness of our results to the exclusion of the items of the scale that perform most poorly.}

@@

After having studied the factor structure underlying the 11 common items in the previous section, we now specify a multiple-group factor analysis model to formally quantify the strength of the relationship between the observed items in our scale and the latent socio-emotional skills, and to test for invariance across cohorts. We specify two groups of children $c=\{BCS,MCS\}$, corresponding to the two cohorts. Each individual child is denoted by $j=1\dots N_c$, where $N_c$ is the number of children in cohort $c$. For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$ corresponding to the eleven maternal reports in \autoref{tab:scalecomp}. Following the EFA results above, we specify two models: one in which we assume that each child is characterised by only one latent skills vector, and another in which we assume that each child is characterised by a latent bi-dimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$.

@@

We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}. We mirror the structure found in the exploratory factor analysis above, and assume that all items load on one factor for the 1-factor solution (\autoref{tab:load_add}), and that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor for the 2-factor solution (\autoref{tab:load}).\footnote{The dedicated factor structure in the two-factor case corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}

@@

Any comparison between socio-emotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts. In other words, the items in our new scale must measure socio-emotional skills in the same way in the BCS and MCS data. This property is denominated measurement invariance (MI) \citep{Vandenberg2000a,Putnick2016}.

@@

In the case where the available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}). Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood. In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$). This makes identification reliant on more stringent normalisations. A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}. New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed. In light of this, they propose models that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts. A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.

@@

In any case, we present a range of fit indices to provide a more complete assessment of measurement invariance. We present the measurement invariance results for the 1-factor model in \autoref{tab:fit_1f}, and those for the 2-factor model in \autoref{tab:fit}. First, by comparing the fit of each nested model across the 1-factor and the 2-factor models, it is clear that the 1-factor model fits the data significantly worse than the 2-factor model, according to all the criteria considered.\footnote{It is  worth noting that threshold and loading invariance only can be established also in the 1-factor case, i.e. intercept invariance is never achieved.} Hence, in our analysis since now on, we adopt the two-factor solution, which is also consistent with the child psychology literature cited above: as mentioned above, we name the two factors externalizing and internalizing skills. We now examine the measurement invariance properties of our chosen two-factor solution in greater details. Looking at Panel A of \autoref{tab:fit}, we see that the overall fit of the configural model for the chosen 2-factor solution is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$. As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels. The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs. In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance. Invariance of loadings and thresholds across cohorts implies that the items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
